Case Study: RNA-Seq Differential Analysis (mir200c knockdown)

### 2018-03-22 datasets/RNA-Seq/RNAseq-gtex.Rmd
Make RNAseq analyses more uniform

```{r, dsd-download}
if (!file.exists(count_file)) {
    download.file("https://www.dropbox.com/s/pptxqnpsmdkyhcf/brain-counts.rds?dl=1",
                  destfile = count_file)
}
dsd <- readRDS(count_file)
```

@@

```{r}
if (!file.exists(result_file)) {
    ## build DESeq object and run DESeq2
    dds <- DESeq(dsd, parallel = TRUE, BPPARAM = multicoreParam)
    res <- results(dds, independentFiltering = FALSE)
    saveRDS(res, file = result_file)
} else {
    res <- readRDS(result_file)
}
```

 We reformat DESeqResuls object into a normal data.frame for further analysis.

```{r}
## build input data.frame
geneExp <- data.frame(geneName = rownames(res),
                      pval = res$pvalue,
                      SE = res$lfcSE, 
                      ind_covariate = res$baseMean,
                      effect_size = res$log2FoldChange,
                      test_statistic = res$stat)
geneExp <- dplyr::as_tibble(geneExp)

## filter data
geneExp <- na.omit(geneExp)
```

@@

In RNA-seq differential expression analysis, it is very well established that the mean expression is an informative covariate that is independent under the null and that can be used to increase power while keeping FDR control (DESeq2 paper).

## Mean Counts

@@

These discreteness, however, corresponded to genes with very low expression values. For example, there were `r nrow( dplyr::filter(geneExp, ind_covariate <= 1) )` genes with less than an average of 1 read across the 20 samples. The distribution of p-values looked much better when removing this very lowly expressed genes.

@@

```{r}
if (!file.exists(bench_file)) {
    bd <- initializeBenchDesign()
    sb <- buildBench(bd, data = geneExp, ftCols = "ind_covariate")
    saveRDS(sb, file = bench_file)
} else {
    sb <- readRDS(bench_file)
}
```

@@

```{r}
assayNames(sb) <- "qvalue"
sb <- addDefaultMetrics(sb)
estimatePerformanceMetrics( sb, tidy=TRUE ) %>%
  filter( performanceMetric == "rejections" ) %>%
  select( blabel, performanceMetric, alpha, value ) %>%
  arrange( desc(value) )
```

>>>>

```{r, dsd-download}
if (!file.exists(count_file)) {
    download.file("https://www.dropbox.com/s/9g06cx9vyonhdao/rnaseq-brain-counts.rds?dl=1",
                  destfile = count_file)
}
dsd <- readRDS(count_file)
```

@@

```{r}
if (file.exists(result_file)) {
    res <- readRDS(result_file)
} else {
    dds <- DESeq(dsd, parallel = TRUE, BPPARAM = multicoreParam)
    res <- results(dds, independentFiltering = FALSE)
    res <- results(dds, independentFiltering = FALSE) %>% 
        as.data.frame() %>%
        na.omit() %>% 
        dplyr::select(pvalue, baseMean, log2FoldChange, lfcSE, stat) %>%
        dplyr::rename(pval = pvalue,
                      ind_covariate = baseMean, 
                      effect_size = log2FoldChange,
                      SE = lfcSE, 
                      test_statistic = stat)
    saveRDS(res, file = result_file)
}
```

@@

In RNA-seq differential expression analysis, it is very well established that the mean expression is an informative covariate that is independent under the null and that can be used to increase power while keeping FDR control.

### Mean Counts

@@

These discreteness, however, corresponded to genes with very low expression values. For example, there were `r nrow(dplyr::filter(geneExp, ind_covariate <= 1))` genes with less than an average of 1 read across the 20 samples. The distribution of p-values looked much better when removing this very lowly expressed genes.

@@

```{r}
if (file.exists(bench_file)) {
    sb <- readRDS(bench_file)
} else {
    bd <- initializeBenchDesign()
    sb <- buildBench(bd, data = geneExp, ftCols = "ind_covariate")
    saveRDS(sb, file = bench_file)
}
```

@@

```{r}
assayNames(sb) <- "qvalue"
sb <- addDefaultMetrics(sb)
estimatePerformanceMetrics(sb, tidy = TRUE) %>%
    filter(performanceMetric == "rejections") %>%
    select(blabel, performanceMetric, alpha, value) %>%
    arrange(desc(value))
```

### 2018-03-22 datasets/RNA-Seq/RNAseq-mir200c.Rmd
Make RNAseq analyses more uniform

As a second RNA-seq dataset, we will test for differences in gene expression upon the knockout of the microRNA mir-200c [@Kim_2013].  The raw fastq files can be found under the accession number `SRP030475`. As the number of samples is limited the experiment might be underpowered, as in most RNA-seq analysis. This is an experimental scenario that could benefit from power gaining from using modern FDR control methods.

@@

## intermediary files we create below
count_path <- file.path(datdir, "SRP030475")
count_file <- file.path(count_path, 'rse_gene.Rdata')
bench_file <- file.path(resdir, "mir200c-benchmark.rds")

@@

```{r}
if( !file.exists( count_file ) ){
    download_study('SRP030475', outdir=count_path)
}

load(count_file)
dsd <- scale_counts(rse_gene)
```

@@

```{r}
dsd <- dsd[,grepl("WT|200c", colData(dsd)$title)]
colData(dsd)$mir200c <- factor(ifelse( grepl( "WT", colData(dsd)$title), "WT", "KO"))
dsd <- as( dsd, "DESeqDataSet" )
storage.mode(assays(dsd)[["counts"]]) <- "integer"
```

@@

## Differential testing

@@

```{r}
design(dsd) <- ~ mir200c
dsd <- DESeq(dsd)
res <- results(dsd, independentFiltering=FALSE) %>% 
  as.data.frame(  ) %>%
  na.omit() %>% 
  dplyr::select( pvalue, baseMean, log2FoldChange, lfcSE, stat ) %>%
  dplyr::rename( pval=pvalue, ind_covariate=baseMean, 
          effect_size=log2FoldChange, SE=lfcSE, 
          test_statistic=stat ) 
head( res )
```

## Mean Counts

As with the previous example, the mean counts is an informative covariate.

```{r, meancov-diag-scatter, fig.width=4.5, fig.height=3.5}
rank_scatter( res, pvalue = "pval", covariate = "ind_covariate") +
    ggtitle("Mean coverage as independent covariate") +
    xlab("Mean Expression")
```

Similarly to the first dataset, keeping all the tests results in a strange discretness. This is removed once we filter very lowly expressed genes. For the first covariate bin, however, there is a strange behaviour in which the distribution seems a bit skewed towards larger p-values.

```{r, meancov-diag-hist, fig.width=10, fig.height=3.2}
strat_hist( res, pvalue = "pval",
           covariate = "ind_covariate", maxy = 7.5)
res <- filter(res, ind_covariate > 1 )
strat_hist( res, pvalue = "pval",
           covariate = "ind_covariate", maxy = 3)
```

@@

```{r}
if (!file.exists(bench_file)) {
    bd <- initializeBenchDesign(nmids=10000)
    sb <- buildBench(bd, data = res, ftCols = "ind_covariate")
    saveRDS(sb, file = bench_file)
} else {
    sb <- readRDS(bench_file)
}
```

@@

```{r}
head( assays( sb )[["bench"]] )
```

@@

```{r}
assayNames(sb) <- "qvalue"
sb <- addDefaultMetrics(sb)
estimatePerformanceMetrics( sb, tidy=TRUE ) %>%
  filter( performanceMetric == "rejections", alpha == 0.1 ) %>%
  select( blabel, performanceMetric, alpha, value ) %>%
  arrange( desc(value) )
```

>>>>

As a second RNA-seq dataset, we will test for differences in gene expression upon the knockout of the microRNA mir-200c [(Kim et al., 2013)](http://dx.doi.org/10.1038/nsmb.2701).  The raw fastq files can be found under the accession number `SRP030475`. As the number of samples is limited the experiment might be underpowered, as in most RNA-seq analysis. This is an experimental scenario that could benefit from power gaining from using modern FDR control methods.

@@

## intermediary files we create below
count_file <- file.path(datdir, "rse_gene.Rdata")
result_file <- file.path(resdir, "mir200c-results.rds")
bench_file <- file.path(resdir, "mir200c-benchmark.rds")

@@

```{r}
if (!file.exists(count_file)) {
    download_study('SRP030475', outdir = datdir)
}
load(count_file)
dsd <- scale_counts(rse_gene)
```

@@

```{r}
dsd <- dsd[, grepl("WT|200c", colData(dsd)$title)]
colData(dsd)$mir200c <- factor(ifelse(grepl("WT", colData(dsd)$title), "WT", "KO"))
dsd <- as(dsd, "DESeqDataSet")
storage.mode(assays(dsd)[["counts"]]) <- "integer"
```

@@

## Differential Testing

@@

```{r}
if (file.exists(result_file)) {
    res <- readRDS(result_file)
} else {
    design(dsd) <- ~ mir200c
    dds <- DESeq(dsd)
    res <- results(dds, independentFiltering = FALSE) %>% 
        as.data.frame() %>%
        na.omit() %>% 
        dplyr::select(pvalue, baseMean, log2FoldChange, lfcSE, stat) %>%
        dplyr::rename(pval = pvalue,
                      ind_covariate = baseMean, 
                      effect_size = log2FoldChange,
                      SE = lfcSE, 
                      test_statistic = stat)
    saveRDS(res, file = result_file)
}
```

## Covariate Diagnostics

As with the GTEx example, the mean counts is an informative covariate.

### Mean Counts

```{r, meancov-diag-scatter, fig.width=4.5, fig.height=3.5}
rank_scatter(res, pvalue = "pval", covariate = "ind_covariate") +
    ggtitle("Mean coverage as independent covariate") +
    xlab("Mean Expression")
```

Similarly to the GTEx dataset, keeping all the tests results in a strange discretness. This is removed once we filter very lowly expressed genes. For the first covariate bin, however, there is a strange behaviour in which the distribution seems a bit skewed towards larger p-values.

```{r, meancov-diag-hist, fig.width=10, fig.height=3.2}
strat_hist(res, pvalue = "pval",
           covariate = "ind_covariate", maxy = 7.5)
res <- filter(res, ind_covariate > 1)
strat_hist(res, pvalue = "pval",
           covariate = "ind_covariate", maxy = 3)
```

@@

```{r}
if (file.exists(bench_file)) {
    sb <- readRDS(bench_file)
} else {
    bd <- initializeBenchDesign(nmids = 10000)
    sb <- buildBench(bd, data = res, ftCols = "ind_covariate")
    saveRDS(sb, file = bench_file)
}
```

@@

```{r}
head(assays(sb)[["bench"]])
```

@@

```{r}
assayNames(sb) <- "qvalue"
sb <- addDefaultMetrics(sb)
estimatePerformanceMetrics(sb, tidy = TRUE) %>%
    filter(performanceMetric == "rejections", alpha == 0.1) %>%
    select(blabel, performanceMetric, alpha, value) %>%
    arrange(desc(value))
```
